sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
- add main equation - remove ELPD description (could add back in a Draper ref for log-scoring) - add description for 1st example

TODO ADD EQUATIONS

@@

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Figure \ref{fig:diagram}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue [@rufener2021] is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model.
Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

@@

As a second example, we will construct a spatiotemporal model of catch rates of
Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
coast of Vancouver Island, Canada. This example extends the spatial model
described above by also including (1) spatiotemporal fields allowing the
distribution in each year to be unique, (2) a time-varying intercept as an
AR(1) process to allow each year to have a different mean catch rate, (3)
a smooth effect of depth, allowing catch rates to vary non-linearly with
depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
dependent. Since catch rates are positive, continuous, and contain zeros, we begin
by specifying the response family as a Tweedie distribution [@tweedie1984] with
a log link. We then compare alternative families, spatiotemporal random field
structures, and the inclusion of anisotropy to illustrate the flexibility of
\pkg{sdmTMB}.

>>>>

@@

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\boldsymbol{\beta}$ and offset $O_{\boldsymbol{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\boldsymbol{s}}$, spatiotemporal intercept random fields $\epsilon_{\boldsymbol{s},t}$, time-varying effects $\boldsymbol{\gamma_t}$, and spatially varying effects $\zeta_{\boldsymbol{s}}$ (Figure \ref{fig:diagram}). These components are discussed in more detail via our examples below.

$$
\begin{aligned}
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
O_{\boldsymbol{s},t} +
\alpha_g +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} + 
\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \boldsymbol{\gamma_t} +
\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} \right),
\end{aligned}
$$
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue [@rufener2021] is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model.
Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

where $\boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\boldsymbol{\beta}$ represents estimated main effect coefficients, and $\omega_{\boldsymbol{s}}$ represents the estimated spatial field,

@@

As a second example, we will construct a spatiotemporal model of catch rates of
Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
coast of Vancouver Island, Canada. This example extends the spatial model
described above by also including (1) spatiotemporal fields allowing the
distribution in each year to be unique, (2) a time-varying intercept as an
AR(1) process to allow year effects to vary but remain autocorrelated, (3)
a smooth effect of depth, allowing catch rates to vary non-linearly with
depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
dependent. Since catch rates are positive, continuous, and contain zeros, we begin
by specifying the response family as a Tweedie distribution [@tweedie1984] with
a log link. We then compare alternative families, spatiotemporal random field
structures, and the inclusion of anisotropy to illustrate the flexibility of
\pkg{sdmTMB}.
